Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equations

### 2014-04-10 pituba/article/article.tex
more and more

We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium.
For this, we assume a constant isothermal compressibility to oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase. This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the pure effect of gas hydrostatic pressure on its density.

@@

% Gas
\begin{equation}\label{eq: Sg5}
V_{p2}\left(\frac{\bar{S}_{g2}}{B_{g2}} +\frac{\bar{S}_{o2} R_{s2}}{B_{o2}}\right) = N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right) +\left(G_i-G_p\right) \, ,
\end{equation}
%
\begin{equation}\label{eq: S15}
S_o+S_w+S_g=1 \, .
\end{equation}

Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil measured in reservoir conditions.

@@

A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}

>>>>

We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium.
For this, we assume a constant isothermal compressibility for oil phase
\begin{equation}\label{eq: grado}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
\end{equation}
and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}
\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase. This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure on its density.

@@

% Gas
\begin{equation}\label{eq: Sg5}
V_{p2}\left(\frac{\bar{S}_{g2}}{B_{g2}} +\frac{\bar{S}_{o2} R_{s2}}{B_{o2}}\right) = N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right) +\left(G_i-G_p\right) \, ,
\end{equation}
%
\begin{equation}\label{eq: S15}
\bar{S}_o+\bar{S}_w+\bar{S}_g=1 \, .
\end{equation}

Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir conditions.

@@

A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}
